---
title: "Additional analyses"
author: Derek Powell & Zachary Horne
output: 
  html_notebook: 
    code_folding: hide
---

This is a summary of the fifth study mentioned in footnote 1.

## Method

Methods were essentially identical to those of study 2A, with the exception of the actual items that were used.

### Items

| Item         | High-probability event                                       | Low-probability event                                        |
| ------------ | ------------------------------------------------------------ | ------------------------------------------------------------ |
| carFactory   | A factory worker is crushed to death due to an accident at an automobile plant in India. | A factory worker is crushed to death due to an accident at an automobile plant in Japan. |
| windowFall   | A window washer is killed instantly when he falls from the 10th floor after a cable snaps. | A window washer is killed instantly when he falls from the 2nd floor after a cable snaps. |
| highRise     | 200 people die when a high-rise apartment building in China collapses. | 200 people die when a high-rise apartment building in South Korea collapses. |
| breastCancer | A 60 year-old woman dies from breast cancer.                 | A 60 year-old man dies from breast cancer.                   |
| execution    | A man in Iran is executed after being falsely convicted of a crime and sentenced to death. | A man in Japan is executed after being falsely convicted of a crime and sentenced to death. |
| drunkDriver  | A pedestrian is killed by a drunk driver on a Friday night   | A pedestrian is killed by a drunk driver on a Friday morning |
| ankle        | An 8 year-old boy breaks his ankle while playing soccer.     | An 8 year-old boy breaks his ankle while playing tennis.     |
| allergic     | An adolescent boy is hospitalized after having his first sudden allergic reaction to peanuts. | An adolescent boy is hospitalized after having his first sudden allergic reaction to potatoes. |
| ebola        | 300 people die in Liberia after being exposed to Ebola.      | 300 people die in Egypt after being exposed to Ebola.        |

## Results

Below is a figure summarizing the results of this study.

```{r}
library(tidyverse)

stderror <- function(x) {
          sqrt(var(x[!is.na(x)]) / length(x[!is.na(x)]))
}

outLik <- read.csv("Data/wiw-outcome-likert-R.csv", skip = 0, header = T)

outLik <- outLik %>% filter(DistributionChannel!="preview")
# remove unused qualtrics variables
remove_cols <- c("RecipientLastdata",
                 "RecipientFirstdata",
                 "RecipientEmail",
                 "Finished",
                 # "ResponseId",
                 "ExternalReference",
                 "DistributionChannel",
                 "UserLanguage",
                 "Status")

outLik <- outLik[ , -which(colnames(outLik) %in% remove_cols)]
outLik <- outLik %>%
  filter(check5==5, check1==1) %>%
  gather(key=item, value = response, carFactory:fireFighter)



outLik <- outLik %>% mutate(response3=ifelse(response>3,1,ifelse(response < 3,-1, 0)))

outLik <- outLik %>% mutate(response3r = response3*-1)


```

```{r}

outLik %>% 
  group_by(item) %>% 
  # filter(condition=="worse") %>%
  summarize(mean=mean(response3r), 
            ll = mean(response3r) - stderror(response3r), 
            ul = mean(response3r) + stderror(response3r)) %>% 
  ggplot(aes(x=reorder(item,mean), y=mean, ymin=ll, ymax=ul)) + 
  geom_pointrange() + 
  geom_hline(yintercept=0) +
  labs(x="Item", y="Average Response")

```

_Figure:_ Average responses by item for Study 5. Error bars indicate standard errors. Responses are represented using scale-means for visualization purposes only (higher scores indicate greater bias toward unexpected event).